---
author: "Philipp Satlawa - h0640348"
date: "07/02/2021"
title: "Statistics with R - Exercise 3"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document contains the answered questions of exercise 3 for the course "Statistics with R".

***

## Task 1 – Statistical Tests
Load packages
```{r}
library(nortest)
library(exactRankTests)
library(car)
library(ggplot2)
```

### 1. One Sample Test to test H0 - if the mean length of the sample is 950mm
```{r}
# create vector with sample of size n = 18 from machine nr 1
(x <- c(870, 930, 932, 935, 938, 1045, 1050, 1052, 1055,
       970, 980, 1001, 1009, 1027, 1030, 1032, 1040, 1046))

# check if the sample has normal distribution
lillie.test(x = x)
```
The Lilliefors normality tests p-value is (0.04) below 0.05, hence have to reject the H0 stating that the sample has a normal distribution. Thus we cannot use the one-sample t-test, since it requires a normal distribution of the data. We will use two non parametric tests instead.

**Wilcoxon Signed Rank Test**
```{r}
# applying the Wilcoxon Signed Rank Test
wilcox.exact(x = x, alternative = "two.sided", mu = 950)
```
The p-value of 0.00321 is lower than 0.05, hence we reject H0 stating that the mean length of the produced components is 950mm.

**Sign-Test**
```{r}
# reference mean value
M0 <- 950
# calculate differences to reference
(diff_lenghts <- x - M0)
# create empty vector z 
z <- numeric(length = length(diff_lenghts))
# set values of vector z to 0, if differences to reference are smaller than 0
z[diff_lenghts < 0] <- 0
# set 1 otherwise
z[diff_lenghts > 0] <- 1
# two sided binomial test with x = number of ones in vector z
binom.test(x = sum(z), n = length(diff_lenghts), p = 0.5, alternative = "two.sided")
```
As the p-value is larger than 0.05, there is no evidence in the data to reject H0, hence we can assume that the mean length is 950mm. However one requirement for the Sign-test is data symmetricity and cannot assume this for our data (please see 4. Informative plots - density plot).


### 2. Two Sample Test to test H0 - if the variance of both machines product lengths is the same
```{r}
# create vector with sample of size n = 18 from machine nr 2
y <- c(1025, 1006, 1005, 1000, 990, 1013, 912, 1011, 909, 947)
# normality test for new machine
lillie.test(x = y)
```
Since the p-value (0.03) is less than 0.05 the sample is not passing the test for normal distribution, similarly to the data of the first machine.
Because the data is not normally distributed we cannot use the F-Test and we will instead use the Levene Test.

**Levene Test**
```{r}
# combine the two vectors with the sample data into a dataframe
data_frame <- data.frame(len = c(x, y),
                         group = c(rep("old machine", length(x)), rep("new machine", length(y))),
                         rank = rank(c(x, y)))

# perform Levene Test
leveneTest(len ~ group, data = data_frame, center = "mean")
```
The p-value of the Levene Test is 0.262 (> 0.05), thus we cannot reject H0 stating that the varinaces of the two samples are equal.

### 3. Two Sample Test to test H0 - if the mean length is the same with the new machine
Since we know that the data is not normally distributed, we need to perform a non-parametric test in our case the Wilcoxon Rank Sum Test to test for the H0 on the two samples.
```{r}
# Wilcoxon Rank Sum Test, 2-sided and equal variances
wilcox.test(len ~ group, data = data_frame, alternative = "two.sided", exact = TRUE)
```
We obtain a p-value of 0.226, thus we do not reject H0 stating that the equal mean length of both machines does not differ.

### 4. Informative plots
```{r}
# boxplot of product lengths produced by the two machines
boxplot(len ~ group, data = data_frame,
        col = heat.colors(2),
        main = "Product lengths",
        names = c("old machine", "new machine"), ylab = "Length [mm]")

# density plot of product lengths produced by the two machines with the target length = 950mm
ggplot(data_frame, aes(x = len)) + geom_density(aes(color = group)) + geom_vline(aes(xintercept = M0), size=1, color="blue") + labs(x= "Length [mm]", subtitle="Product length distribution of both machines and target length") + theme(legend.position="bottom")

```

***

## Task 2 - Functions

### 1. Create a function with the name quad_equ()
```{r}
quad_equ <- function(coef){
  x1 <- -(coef[1]/2) - sqrt((coef[1]/2)^2-coef[2])
  x2 <- -(coef[1]/2) + sqrt((coef[1]/2)^2-coef[2])
    
  return(c(x1,x2))
}
coef <- c(3,-4)
quad_equ(coef)
```


### 2. Extend function - check if the passed argument coef gives real solutions
```{r}
quad_equ <- function(coef){
  if((coef[1]/2)^2-coef[2] >= 0){
    x1 <- -(coef[1]/2) - sqrt((coef[1]/2)^2-coef[2])
    x2 <- -(coef[1]/2) + sqrt((coef[1]/2)^2-coef[2])
    
    return(c(x1,x2))
  } else{
    print("The supplied vector x has no real solution.")
    #stop("The supplied vector x has no real solution.")
  }
}
coef <- c(3,-4)
quad_equ(coef)
coef <- c(-2, 2)
quad_equ(coef)
```

### 3. Extend function - check if the passed argument is valid 
```{r}
quad_equ <- function(coef){
  if(length(coef)==2){
    if((coef[1]/2)^2-coef[2] >= 0){
      x1 <- -(coef[1]/2) - sqrt((coef[1]/2)^2-coef[2])
      x2 <- -(coef[1]/2) + sqrt((coef[1]/2)^2-coef[2])
      
      return(c(x1,x2))
    } else{
      print("The supplied vector x has no real solution.")
    }
  } else{
    print("The supplied vector x has the wrong length. 
         The input vector has to have length 2. 
         Please enter a valid vector.")
  }
}
# test if the function is working correctly with a correct input
coef <- c(3,-4)
quad_equ(coef)
# test if the function invokes an error with an incorrect input
coef <- c(3,-4, 9)
quad_equ(coef)
```

***

## Task 3 - Graphics

Creation of a statistical graph

# Load libraries
```{r}
library(data.table)
```

### 1. Create a data for the construction of a plot that mimics the template "Plot_9.png"
```{r}
# create data
dt <- data.table(id = c(1:29), age = c(25, 21, 5, 15, 47, 33, 39, 56, 3, 45, 31, 28, 44, 15, 13, 22, 40, 39, 59, 13, 37, 14, 60, 44, 25, 21, 35, 33, 29))
```

### 2. combine the feature "age" into age intervals 
```{r}
# create vector containing the cutting points
cut_points <- c(0,11,21,31,41,51,61)
# create vector containing the labels
labels <- c("0-10","11-20", "21-30", "31-40", "41-50", "51-60")
# categorize the feature "age" into the created labels 
age_cat <- cut(dt$age, breaks=cut_points, include.lowest=TRUE, right=FALSE, labels=labels)
# add age_cat to data.table
dt[, ("age_category") := age_cat]
```

### 3. Create plot
```{r}
# Create plot
p <- ggplot(data = dt, mapping = aes(x=age_category)) + 
  geom_bar(stat="count", fill="orange", color="black", width = 1) +
  labs(title = "Number of People in Each Age Category", x = "Age Category", y = "People") +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid.major = element_line(colour = "grey85"),
        panel.grid.major.x = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_text(colour = "grey30"),
        plot.title = element_text(colour = "grey30")) +
  scale_y_continuous(breaks = seq(0, 9, 1), expand=c(0,0), limits=c(0,9))
p
```

***